Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance

### 2019-06-26 Manuscript/manuscript.Rmd
edits

We then investigated the proportion of class imbalance, defined as the ratio of TAD boundary regions versus the sum of both TAD and non-TAD boundary regions. We compared the estimated the number of TAD boundary regions ($Y=1$ in the response vector) and the rest of genomic bins ($Y=0$) for each cell line, at each of the four resolutions of Hi-C data. The proportional imbalance was found to be invariant to resolution, remaining consistently low for both cell lines, with very little variability between chromosomes (Figure 3C).

@@

We created an ensemble framework to systematically compare different data driven characteristics to identify the optimal set of parameters when predicting TAD boundaries using functional genomic annotations (Figure 3). Each system in the framework was assumed to be composed of a random forest model constructed from a particular cell line, $L = \{GM12878, K562\}$. Within each cell line we evaluated 4 data resolutions, $R = \{10 kb, 25 kb, 50 kb, 100 kb\}$. We then considered 3 separate predictor types as the feature space including $P = \{OC, OP, Distance\}$. To assess the severity of class imbalance, 4 re-sampling techniques were implemented prior to modelling, including $B = \{None, RUS, ROS, SMOTE\}$. Each combination of the above parameters was implemented on chromosome specific sections of the linear genome. Performances were then aggregated across each of the 21 chromosomes (omitting chromosome 9). Letting $|X|$ denote the cardinality of each set, we evaluated $|L| \times |R| \times |P| \times |B|$ different ensemble systems, totaling 2016 ($2 \times 4 \times 3 \times 4 \times 21$) separate models. We report our main results for the GM12878 cell line and have included results from the K562 cell line in the supplementary materials.

![**Figure 3. Ensemble Framework.** A diagram of the cell line specific ensemble framework detailing the combination of inputs used in the model building pipeline. The minoirty and majority classes were represented by the genomic bins that overlapped with flanked TAD boundaries or did not, respectively. Functional genomic annotations obtained from the UCSC Genome Browser Database composed the feature space ($\textbf{X} = \{X_1, X_2, ..., X_p\}$). A feature engineering step was included to isolate the predictor-type specific spatial relationship between annotations and genomic bins in increasing granularity, denoted by the red gradient. The data was split into a 7:3 training to testing set ratio. Data re-sampling was then applied to the training set. Lastly, a random forest classification algorithm was implemeted and validated on the testing data.](Figures/figure3_ensemble_framework.png)

>>>>

We then investigated the proportion of class imbalance, defined as the ratio of TAD boundary regions versus the sum of both TAD and non-TAD boundary regions. We compared the estimated the number of TAD boundary regions ($Y=1$ in the response vector) and the rest of genomic bins ($Y=0$) for each cell line, at each of the four resolutions of Hi-C data. The proportional imbalance was found to be invariant to resolution, remaining consistently low for both cell lines, with very little variability between chromosomes (Figure 3C). Such low proportions of imbalance gave precedence for the implementation of re-sampling prior to modeling.

@@

In order to identify the optimal set of parameters necessary for TAD boundary prediction, we created an ensemble framework to systematically compare performances of random forest classifiers built on different combinations of predictor types and re-sampling techniques (Figure 4). Each system in the framework was assumed to be composed of a random forest model constructed from a particular cell line ($L = \{GM12878, K562\}$). Within each cell line we evaluated 4 data resolutions ($R = \{10 kb, 25 kb, 50 kb, 100 kb\}$). We then considered 3 separate predictor types as the feature space ($P = \{OC, OP, Distance\}$. To assess the severity of class imbalance, 4 re-sampling techniques were implemented prior to modelling, including $B = \{None, RUS, ROS, SMOTE\}$). Each combination of the above parameters was implemented on chromosome specific sections of the linear genome. Performances were then aggregated across each of the 21 chromosomes (omitting chromosome 9). Letting $|X|$ denote the cardinality of each set, we evaluated $|L| \times |R| \times |P| \times |B|$ different ensemble systems, totaling 2016 ($2 \times 4 \times 3 \times 4 \times 21$) separate models. <!--We report our main results for the GM12878 cell line and have included results from the K562 cell line in the supplementary materials.-->

![**Ensemble Framework.** A diagram of the ensemble framework detailing the combination of inputs used in the model building pipeline. TAD boundary data was extracted from a specific Hi-C resolution and a variety of functional genomic annotations were used to establish the feature space. Each subsequent cell line specific random forest classification model was built on an isolated predictor type in conjunction with a particular data re-sampling technique to account for class imbalance. The model was then validated on a testing set and performances were evaluated using MCC.](Figures/figure3_ensemble_framework_V2.png)
